---
title: "semver: Basics"
author: "John D Harrison"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{semver: Basics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The goal of this vignette is to describe the basic functionality of the 
`semver` package.

## Introduction

The `semver` package provides tools and functions for parsing, rendering 
and operating on semantic version strings. Semantic versioning is a simple 
set of rules and requirements that dictate how version numbers are 
assigned and incremented as outlined at [http://semver.org](http://semver.org).

A basic summary of how semantic versioning operates is given a version number MAJOR.MINOR.PATCH, increment the:

1. MAJOR version when you make incompatible API changes,
2. MINOR version when you add functionality in a backwards-compatible manner, and
3. PATCH version when you make backwards-compatible bug fixes.

Additional labels for pre-release and build metadata are available as extensions to the MAJOR.MINOR.PATCH format.

## `semver` package

The `semver` package provides a wrapper for the C++14 semantic versioning parser
written by Marko Živanović. The project is currently hosted on 
[github](https://github.com/zmarko/semver). The [Rcpp package](https://github.com/RcppCore/Rcpp) was used to provide R bindings. Some 
changes were made on the C++ side as currently CRAN does not accept packages
compiling under C++14 (R version 3.4.0 should allow this in future).

### Parsing/Rendering semantic versions

#### parse_version
The `parse_version` function parses a character vector containing valid 
semantic versioning strings returning an "svlist" object. 

```{r example1}
library(semver)
examples <- c("1.0.0", "2.1.3", "1.0.0-alpha", "1.0.0-alpha+1.2", 
              "1.8.2-beta.1.13", "1.8.2-beta.1.10")
sem_versions <- parse_version(examples)
sem_versions
str(sem_versions)
```

#### render_version
The `render_version` function acts on an "svptr"/"svlist" object. It 
returns an R list/(list of lists) giving the major, minor and patch 
version as an integer and the prerelease and build version as a charcter 

```{r example1a}
render_version(sem_versions[c(1, 4)])
render_version(sem_versions[[5]])
str(render_version(sem_versions[[5]]))
```

### Comparing versions

The `parse_version` function returns a list of `svptr` objects. These `svptr`
objects represent the semantic versions. We can do comparisons like:

#### svptr Ops

```{r example2}
sem_versions[[1]] <= sem_versions[[5]]
sem_versions[[1]] > sem_versions[[5]]

# compare example 5, 6 (pre-release ordering matters)
sem_versions[[5]] > sem_versions[[6]]

# compare example 3, 4 (build order does not matter)
sem_versions[[3]] == sem_versions[[4]]
```

#### Summary of svlist objects

You can get the min, max and range of the versions

```{r example3}
min(sem_versions)
max(sem_versions)
range(sem_versions)
```

#### Sort, Order and rank an svlist
You can sort, order and rank the versions:
```{r example4}
sort(sem_versions)
order(sem_versions)
rank(sem_versions)
```

#### Ops on svlist

You can also compare "svlist" objects. If the lengths of the two lists are
unequal recycling occurs:

```{r example4a}
sem_versions > sem_versions[1]

```

### Compare to character strings

Sometimes it can be useful to compare a parsed vector of semantic versions to a 
character string:

```{r example5}
sem_versions > "1.1.0-beta"
sem_versions[sem_versions > "1.1.0-beta"]
```

## Updating versions

If you want to change or ammend the major, minor, patch, prerelease or build
fields the semver package provides a number of methods to achieve this. 

### set_version

The `set_version` method operates on `svptr` and `svlist` classes. It
simply changes the indicated field to the value given. Other fields in the
version are unaffected.

#### svptr

For the `svptr` class the `set_version` method takes a character string 
argument `field` which indicates which field (major, minor etc.) to change. 
The `value` argument is an integer scalar when field is major, minor or 
patch and a character string when prerelease or build. 

```{r example6}
library(semver)
examples <- c("1.0.0", "2.1.3", "1.0.0-alpha", "1.0.0-alpha+1.2", 
              "1.8.2-beta.1.13", "1.8.2-beta.1.10")
sem_versions <- parse_version(examples)
set_version(sem_versions[[1]], "major", 2L)
set_version(sem_versions[[1]], "minor", 1L)
set_version(sem_versions[[1]], "patch", 1L)
set_version(sem_versions[[4]], "prerelease", "beta")
set_version(sem_versions[[4]], "build", "bld1a")

```
**Note that changing a field with the set_version method does not change any other field.**

##### Dollar assignment

As syntactic sugar for assigning using `set_version` there is a dollar assignment method 
for svptr classes so the following are equivalent waysto assign fields:

```{r example7}
sem_versions[[1]] <- set_version(sem_versions[[1]], "major", 3L)
sem_versions[[1]]

# Syntactic sugar
sem_versions[[1]]$minor <- 2L
sem_versions[[1]]

```

#### svlist

For `svlist` classes the `set_version` method expects a character vector
for the `field` argument. The `value` argument is expected to be a list
with integer and character elements assigning to (major, minor, patch)/(prerelease, build)
fields respectively. If either the length of the field or values argeument
is shorter than the number of elements in the `svlist` then recycling occurs.

```{r example8}
examples <- c("1.0.0", "1.8.2-beta.1.10", "2.4.6-8")
sem_versions <- parse_version(examples)
# recycling on the field argument
set_version(sem_versions, "major", list(2L, 4L, 6L))
# recycling on the value argument
set_version(sem_versions, c("major", "minor", "patch"), list(7L))
# assigning integer and character values
set_version(sem_versions, c("prerelease", "minor", "build"), 
            list("alpha", 3L, "build1.12"))
```


### reset_version

The `reset_version` method operates on `svptr` and `svlist` classes. It
changes the indicated field to the value given. Fields with lower precedence
are set to default values:

```
MAJOR(0L) > MINOR (0L) > PATCH (0L) > PRERELEASE ("") > BUILD ("")
```

#### svptr
The `reset_version` method for `svptr` classes operates similarly to the
`set_version` method with fields of lower precedence being set to default values:

```{r example9}
examples <- c("1.8.2-beta.1.10+somebuild", "2.4.6-8")
sem_versions <- parse_version(examples)
reset_version(sem_versions[[1]], "major", 2L)
reset_version(sem_versions[[1]], "minor", 3L)
reset_version(sem_versions[[1]], "patch", 4L)
reset_version(sem_versions[[1]], "prerelease", "gamma")
reset_version(sem_versions[[1]], "build", "superbuild")
```

#### svlist
The `reset_version` method for `svlist` classes operates similarly to the
`set_version` method with fields of lower precedence being set to default 
values. Again recycling of elements occur if the length of the field/value
argument is shorter than the number of elements in the `svlist`:

```{r example10}
examples <- c("1.8.2-beta.1.10+somebuild", "2.4.6-8")
sem_versions <- parse_version(examples)
# recycling on both arguments
reset_version(sem_versions, "major", list(3L))
# recycling on field argument
reset_version(sem_versions, "minor", list(3L, 4L))
# recycling on value argument
reset_version(sem_versions, c("major", "patch"), list(4L))
# assigning integer and character fields
reset_version(sem_versions, c("prerelease", "minor"), list("zeta", 7L))

```
### increment_version

The `increment_version` method operates on `svptr` and `svlist` classes. It
increments the given field with the provided value. Fields of lower precedence
are set to default value. 

1. **Only major, minor and patch field can be incremented**

2. **To decrement a field provide a negative integer as the value argument**

#### svptr
The `increment_version` method for `svptr` classes increments the chosen field with
the given value with fields of lower precedence being set to default values:

```{r example11}
examples <- c("1.8.2-beta.1.10+somebuild", "2.4.6-8")
sem_versions <- parse_version(examples)
# incrementing versions
increment_version(sem_versions[[1]], "major", 1L)
increment_version(sem_versions[[1]], "minor", 2L)
increment_version(sem_versions[[1]], "patch", 3L)
# decrementing versions
increment_version(sem_versions[[1]], "major", -1L)
increment_version(sem_versions[[1]], "minor", -2L)
increment_version(sem_versions[[1]], "patch", -2L)
```

#### svlist

The `increment_version` method for `svlist` classes takes a character vector 
as argument `field` and an integer vector as argument `value`. Recycling occurs
as for `set_version`/`reset_version` methods:

```{r example12}
examples <- c("1.8.2-beta.1.10+somebuild", "2.4.6-8")
sem_versions <- parse_version(examples)
## Incrementing
# recycling on both arguments
increment_version(sem_versions, "major", 3L)
# recycling on field argument
increment_version(sem_versions, "minor", c(3L, 4L))
# recycling on value argument
increment_version(sem_versions, c("major", "patch"), 4L)
## Decrementing
# recycling on both arguments
increment_version(sem_versions, "major", -1L)
# recycling on field argument
increment_version(sem_versions, "minor", c(-3L, -4L))
# recycling on value argument
increment_version(sem_versions, c("minor"), -4L)

```
